---
title: "🐟 Acoustic Receiver Array Design and Movement Simulation"
subtitle: "Optimizing Telemetry Systems and Simulating Fish Movement Patterns"
author: 
  - "Jake Brownscombe"
date: "`r Sys.Date()`"
output: 
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    fig_width: 10
    fig_height: 8
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Acoustic Receiver Array Design and Movement Simulation}
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  cache = FALSE,
  fig.width = 10,
  fig.height = 8
)

# Suppress verbose output from positionR functions
options(positionR.verbose = FALSE)
```

<div class="alert alert-info">
<h4>**Analysis Overview**</h4>
This tutorial demonstrates acoustic telemetry array design and movement simulation using the **positionR** package. This comprehensive workflow provides tools for optimizing receiver placement, modeling detection efficiency, and simulating realistic fish movement patterns to evaluate system performance before deployment.

**Key Applications:**
- 📡 Optimize receiver array configurations for maximum coverage
- 📊 Model detection efficiency based on environmental conditions
- 🐟 Simulate realistic fish movements and detection patterns
- 🏡 Estimate space use and home range from simulated data
- 🌿 Analyze habitat selection patterns

Ideal for researchers planning acoustic telemetry studies who want to evaluate different array designs and predict system performance before field deployment.
</div>

---

# 🚀 **Setup & Data Preparation**

## Load Required Libraries

```{r libraries}
library(positionR)  
library(raster)
library(sf)
library(ggplot2)
library(dplyr)
library(patchwork)

# Set seed for reproducible results
set.seed(123)
```

## Load Environmental Data

We'll use the example depth raster of Stoney Lake, Ontario, Canada included in the package:

```{r load-data}
# Load example depth raster
data("depth_raster")
raster::crs(depth_raster) <- "EPSG:32617" # UTM Zone 17N

# View raster properties
depth_raster

# Quick visualization
plot(depth_raster, main = "Study Area: Stoney Lake, Ontario", 
     col = colorRampPalette(c("darkblue", "lightblue"))(50))
```

---

# 📡 **Receiver Array Design**

<div class="alert alert-info">
The first critical step in acoustic telemetry is designing an effective receiver array. The spatial configuration of receivers determines detection coverage, positioning accuracy, and ultimately the quality of movement data collected.
</div>

## Generate Array Configurations

positionR provides three methods for generating receiver stations with different spatial arrangements:

```{r generate-stations, fig.width=10, fig.height=4}
# Regular spacing - specified number of points in systematic pattern
stations_regular <- generate_regular_points(depth_raster, n_points = 100, seed = 123)

# Fixed spacing - points separated by specified distance
stations_spaced <- generate_spaced_points(depth_raster, spacing = 1000, seed = 123)

# Random locations - randomly distributed points
stations_random <- generate_random_points(depth_raster, n_points = 100, seed = 123)

# View station data structure
stations_spaced  # Note: includes depth values for each station

# Visualize different array configurations
plot_points_on_input(depth_raster, stations_regular) 
plot_points_on_input(depth_raster, stations_spaced) 
plot_points_on_input(depth_raster, stations_random) 
```

## Calculate Station Distances

Calculate cost-distance and straight-line distance from each receiver to all raster cells. This is essential for modeling detection efficiency across the study area:

```{r calculate-distances}
# Calculate shortest path distances, straight line distances, and tortuosity
station_distances <- calculate_station_distances(
  raster = depth_raster, 
  receiver_frame = stations_regular, 
  station_col = "station_id",
  max_distance = 30000  # 30 km maximum distance
)

# Visualize distance field for one receiver
ggplot(station_distances %>% filter(station_no == 1), 
       aes(x, y, fill = cost_distance)) +
  geom_raster() +
  scale_fill_viridis_c(option = "magma", name = "Distance (m)") +
  geom_point(data = stations_regular %>% filter(station_id == 1), 
             aes(x, y), col = "green", size = 4, inherit.aes = FALSE) +
  labs(title = "Cost Distance Field from Station 1",
       subtitle = "Green point = station location") +
  theme_minimal() +
  coord_sf()

ggplot(station_distances %>% filter(station_no == 1), 
       aes(x, y, fill = tortuosity)) +
  geom_raster() +
  scale_fill_viridis_c(option = "magma", name = "Tortuosity", limits=c(0.9,1.5)) +
  geom_point(data = stations_regular %>% filter(station_id == 1), 
             aes(x, y), col = "green", size = 4, inherit.aes = FALSE) +
  labs(title = "Tortuosity Field from Station 1",
       subtitle = "Green point = station location") +
  theme_minimal() +
  coord_sf()
```

---

# 📊 **Detection Range Modeling**

<div class="alert alert-info">
Detection range is highly variable and depends on environmental conditions including depth, temperature, substrate type, and ambient noise. Accurate detection models are essential for understanding system performance and interpreting detection patterns.
</div>

## Create Depth-Dependent Detection Model

Model detection efficiency as a function of distance and depth. In practice, use range testing data from your study system. Here we create an example model representative of inland lake conditions:

```{r detection-model, fig.width=6, fig.height=8}
# Create logistic detection efficiency model
# These parameters should be based on empirical range testing
logistic_DE <- create_logistic_curve_depth(
  min_depth = 1,           # Minimum depth (m)
  max_depth = 35,          # Maximum depth (m)
  d50_min_depth = 200,     # 50% detection range at min depth (m)
  d95_min_depth = 400,     # 5% detection range at min depth (m)
  d50_max_depth = 500,     # 50% detection range at max depth (m)
  d95_max_depth = 1000,    # 5% detection range at max depth (m)
  plot = TRUE,              
  return_model = TRUE,
  return_object = TRUE
)

# The fitted model can be used for predictions
summary(logistic_DE$log_model)
```

## Apply Detection Model

Apply the detection range model to predict detection efficiency for each receiver across the study area:

```{r predict-efficiency, fig.width=7, fig.height=5}
# Predict detection efficiency for each receiver-cell combination
station_distances$DE_pred <- stats::predict(
  logistic_DE$log_model, 
  newdata = station_distances %>% 
    rename(dist_m = cost_distance) %>% 
    mutate(depth_m = abs(raster_value)),
  type = "response"
)

# Visualize detection efficiency field for one station
ggplot(station_distances %>% filter(station_no == 10), 
       aes(x, y, fill = DE_pred)) +
  geom_raster() +
  scale_fill_viridis_c(option = "magma", name = "Detection\nEfficiency") +
  labs(title = "Detection Efficiency Field - Station 10",
       subtitle = "Probability of detecting a transmission") +
  theme_minimal() +
  coord_sf()
```

## System-Wide Detection Coverage

Analyze detection performance at the system level to evaluate array design effectiveness:

```{r system-efficiency, fig.width=8, fig.height=6}
# Calculate system-level detection probabilities
# Options: "cumulative" (any receiver), "3_plus" (≥3 receivers), "both"
system_DE <- calculate_detection_system(
  distance_frame = station_distances, 
  receiver_frame = stations_regular, 
  model = logistic_DE$log_model, 
  output_type = "cumulative",  # Fast computation
  plots = TRUE
)

# Key outputs:
# system_DE$data - detection probability data
# system_DE$plots$cumulative - visualization
```

---

# 🐟 **Fish Movement Simulation**

<div class="alert alert-info">
Simulating fish movements allows us to evaluate array performance under known conditions. By generating tracks with known properties, we can assess detection probability, positioning accuracy, and space use estimation methods.
</div>

## Generate Realistic Fish Tracks

Generate fish movements using correlated random walks and simulate detections based on the array and detection model:

```{r simulate-movements, fig.width=9, fig.height=7}
# Set start time for temporal consistency
start_time <- as.POSIXct("2025-07-15 12:00:00", tz = "UTC")

# Simulate fish tracks with detections
fish_simulation <- simulate_fish_tracks(
  raster = depth_raster,
  station_distances = station_distances,  # Contains DE_pred values
  n_paths = 10,                          # Number of fish
  n_steps = 100,                         # Steps per track
  step_length_mean = 50,                 # Average step length (m)
  step_length_sd = 30,                   # Step length variability
  time_step = 60,                        # Time between steps (seconds)
  seed = 123,
  start_time = start_time
  
  # Optional species-specific parameters:
  # species = "Walleye",              # Available: "Walleye", "Smallmouth Bass", "Muskellunge"
  # fish_size_cm = 45,                # Fish length in cm
  # behavioral_states = TRUE,         # Three-state movement model
)

# Examine simulation outputs
#head(fish_simulation$tracks)             # Movement tracks
#head(fish_simulation$station_detections) # Individual receiver detections

# Visualize simulated tracks
raster_df <- as.data.frame(depth_raster, xy = TRUE)
ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue", 
                      na.value = "transparent", name = "Depth (m)") +
  geom_path(data = fish_simulation$tracks, 
            aes(x = x, y = y, group = path_id, color = factor(path_id)), 
            linewidth = 0.8) +
  scale_color_discrete(name = "Fish ID") +
  labs(title = "🐟 Simulated Fish Movements",
       subtitle = "Correlated random walks with detection events") +
  theme_minimal() +
  coord_sf()
```

## Analyze Detection Performance

Visualize tracks with detection information and analyze detection performance:

```{r analyze-results, fig.width=10, fig.height=10}
# Plot tracks with detection information
plot_fish_tracks(
  fish_simulation,    # Simulation outputs
  depth_raster,       # Depth raster
  stations_regular,   # Receiver array
  show_detections = TRUE
)

# Analyze detection performance across the array
detection_performance <- analyze_detection_performance(fish_simulation,
                                                       create_plots = TRUE,
                                                       display_plots = FALSE)


performance_plots <- (detection_performance$plots$by_path + 
                      detection_performance$plots$by_station) /
                     (detection_performance$plots$distribution + 
                      detection_performance$plots$time_series)

performance_plots + 
  plot_annotation(
    title = "Detection Performance Analysis",
    subtitle = "Spatial and temporal patterns in detection probability"
  )

```

---

# 🏡 **Space Use Analysis**

<div class="alert alert-info">
Space use analysis is fundamental to understanding animal ecology. Different methods have different assumptions and biases, particularly when dealing with telemetry data that may have detection gaps.
</div>

## Calculate Space Use Estimates

Calculate space use estimates from the simulated tracks using multiple methods:

```{r space-use-analysis, fig.width=8, fig.height=6}
# Calculate space use estimates using multiple methods
space_use_results <- calculate_space_use(
  track_data = fish_simulation$tracks,
  by_fish = TRUE,
  by_time_period = TRUE,
  time_aggregation = "day",  # Options: "hour", "day", "month", "none"
  methods = c("convex_hull", "bounding_box", "grid_cell_count", 
              "ellipse_95", "constrained_convex_hull", "mcp"),
  grid_resolution = 100  
)

# View results
print(space_use_results$space_use_estimates)

# Compare methods graphically
plot_space_use(space_use_results, plot_type = "comparison")
plot_space_use(space_use_results, plot_type = "by_fish")


#plot outputs for select methods

# Select fish and time period 
fish_select <- 1
time_select <- "2025-07-15"

# Convex hull method
plot_space_use(space_use_results, 
               plot_type = "map",
               track_data = fish_simulation$tracks,
               fish_select = fish_select,
               time_select = time_select,                   
               method_select = "convex_hull",
               background_raster = depth_raster) +
  ggtitle("Convex Hull Space Use")

# Constrained convex hull method (raster-based)
plot_space_use(space_use_results, 
               plot_type = "map",
               track_data = fish_simulation$tracks,
               fish_select = fish_select,
               time_select = time_select,                   
               method_select = "constrained_convex_hull",
               background_raster = depth_raster) +
  ggtitle("Constrained Convex Hull Space Use")

# Grid cell count method
plot_space_use(space_use_results, 
               plot_type = "map",
               track_data = fish_simulation$tracks,
               fish_select = fish_select,
               time_select = time_select,                   
               method_select = "grid_cell_count",
               background_raster = depth_raster) +
  ggtitle("Grid Cell Count Space Use")
```



<div class="alert alert-warning">
<h4>**Method Selection Guide**</h4>
- **Grid cell count**: Most representative of actual space use when tracking is consistent, but biased with detection gaps
- **Constrained convex hull**: More robust to detection gaps, consistent with common home range approaches
- **Convex hull**: Simple geometric approach, may overestimate space use in complex habitats

Choose methods based on your data characteristics and research questions.
</div>

---

# 🌿 **Habitat Selection Analysis**

<div class="alert alert-info">
Habitat selection analysis helps identify environmental factors that influence animal distribution. By comparing used (presence) locations to available (absence) locations, we can understand habitat preferences.
</div>

## Generate Presence-Absence Data

Create datasets for habitat selection analysis:

```{r habitat-data, fig.width=6, fig.height=4}
# Generate presence points within space use areas
track_presence <- sample_points_from_track_grid(
  track_data = fish_simulation$tracks,
  reference_raster = depth_raster,
  n_points = 1000,
  crs = 32617
)

# Generate absence points
track_absence<- sample_points_from_system_de(
  system_DE,
  position_points = track_presence,
  n_points = 1000,
  uniform = TRUE,  #set uniform points (not relative to system DE)
  min_prob_threshold = 0.1, #threshold cutoff - only points where >0.1 system DE
  crs = 32617,
  seed = 123
)

#plotting

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_sf(data = track_presence, aes(color = count),
          alpha = 0.5, size = 1) +
  scale_color_viridis_c(name = "Count", option = "magma") +
  coord_sf() +
  theme_minimal() +
  labs(title = "Track Presence Points")

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_sf(data = track_absence %>% filter(fish_id==1 & time_period_label=="2025-07-15"),
          alpha = 1, size = 0.8, col="red") +
  scale_color_viridis_c(name = "Cumulative\nProbability") +
  theme_minimal()+labs(title="Track Absence Points (Uniform Across System)")




```

## Analyze Depth Selection Patterns

```{r habitat-selection, fig.width=6, fig.height=4}
# Add depth values to presence-absence points
track_presence$depth_m <- raster::extract(depth_raster, track_presence)
track_absence$depth_m <- raster::extract(depth_raster, track_absence)

# Combine presence and absence data
track_presence_absence_points <- rbind(
  track_presence %>%
    select(fish_id, time_period_posix, depth_m) %>%
    mutate(type = "presence"),
  track_absence %>%
    select(fish_id, time_period_posix, depth_m) %>%
    mutate(type = "absence")
)


# Analyze depth selection patterns
plot_depth_selection(track_presence_absence_points, plot_type = "density") + 
  ggtitle("Depth Selection from Simulated Tracks") +
  labs(subtitle = "Blue = presences, Red = absences")

# Compare depth selection by fish
plot_depth_comparison(track_presence_absence_points,
                      comparison_var = "fish_id",
                      plot_type = "boxplot") + 
  ggtitle("Depth Selection by Individual Fish")
```

---

# 📝 **Summary & Conclusions**

<div class="alert alert-success">
<h2>**Key Takeaways**</h2>

This tutorial demonstrated a complete acoustic telemetry study workflow:

1. **Array Design**: Generated receiver arrays with different spatial configurations
2. **Detection Modeling**: Created depth-dependent detection range models
3. **Efficiency Analysis**: Calculated system-level detection probabilities
4. **Movement Simulation**: Generated realistic fish movements with detections
5. **Space Use Analysis**: Estimated home ranges using multiple methods
6. **Habitat Analysis**: Created presence-absence data for habitat selection studies

</div>

<div class="alert alert-warning">
<h3>**💡 Best Practices for Array Design**</h3>

- **Detection range models** should be based on empirical range testing data from your study system
- **Array configuration** significantly affects detection probability and positioning accuracy
- **System-level analysis** helps identify coverage gaps and optimize receiver placement
- **Simulation approaches** allow testing of different array designs before costly field deployment
- **Iterative refinement** - use initial deployment data to improve models and optimize array expansion
- **Consider environmental variability** - detection ranges vary with temperature, noise, and other factors

**Next Steps:**
- Conduct range testing in your study area to parameterize detection models
- Use WADE positioning methods for fine-scale movement analysis (see WADE tutorials)
- Compare multiple array configurations to find optimal design for your objectives
</div>
