---
title: "Particle Filter Positioning for Acoustic Telemetry"
subtitle: "Sequential Monte Carlo positioning with movement path estimation"
author:
  - "Jake Brownscombe"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    fig_width: 10
    fig_height: 8
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Particle Filter Positioning}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = 'center',
  cache = FALSE
)
```

<div class="alert alert-info">
<h4>**Analysis Overview**</h4>
This tutorial demonstrates **particle filter positioning** for acoustic telemetry using the positionR package. The particle filter estimates continuous movement paths by propagating a cloud of possible fish locations forward in time, weighting them by detection likelihood at each time step. This approach complements the WADE grid-based method by producing fine-scale path estimates with uncertainty quantification. The tutorial covers the forward filter, two-filter smoother, utilisation distributions, home range estimation, and detection gap analysis.
</div>

# Setup and Libraries

```{r libraries}
library(positionR)
library(raster)
library(sf)
library(ggplot2)
library(ggnewscale)
library(dplyr)
library(patchwork)

set.seed(123)
```

# Environment Setup

```{r env-data}
data("depth_raster")
raster::crs(depth_raster) <- "EPSG:32617"
```

## Receiver Array

```{r receiver-array}
points_regular <- generate_exact_regular_points(depth_raster, n_points = 80, seed = 123)
```

## Station Distances

```{r station-distances}
station_distances <- calculate_station_distances(
  raster = depth_raster,
  receiver_frame = points_regular,
  max_distance = 30000,
  station_col = "station_id"
)
```

## Detection Efficiency Model

```{r detection-model}
logistic_DE <- create_logistic_curve_depth(
  min_depth = 1, max_depth = 35,
  d50_min_depth = 400, d95_min_depth = 800,
  d50_max_depth = 750, d95_max_depth = 1500,
  plot = TRUE, return_model = TRUE, return_object = TRUE
)

station_distances$DE_pred <- stats::predict(
  logistic_DE$log_model,
  newdata = station_distances %>%
    dplyr::rename(dist_m = cost_distance) %>%
    dplyr::mutate(depth_m = abs(raster_value)),
  type = "response"
)
```

# Simulate Fish Tracks

```{r fish-simulation}
start_time <- as.POSIXct("2025-07-15 12:00:00", tz = "UTC")

fish_simulation <- simulate_fish_tracks(
  raster = depth_raster,
  station_distances = station_distances,
  n_paths = 6,
  n_steps = 480,
  step_length_mean = 50,
  step_length_sd = 30,
  time_step = 180,
  seed = 42,
  start_time = start_time,
  include_barriers = TRUE
)
```

# Forward Particle Filter

The particle filter propagates a cloud of particles (possible fish locations) forward in time using a Correlated Random Walk, then weights each particle by how well it explains the observed detections and non-detections.

```{r particle-filter}
pf_results <- particle_filter_positioning(
  detection_data = fish_simulation$station_detections,
  station_info = points_regular,
  de_model = logistic_DE$log_model,
  raster = depth_raster,
  n_particles = 1000,
  step_length_mean = 50,
  step_length_sd = 30,
  turning_angle_sd = 45,
  time_step = 180,
  max_distance = 30000,
  fish_id_col = "path_id",
  time_col = "datetime",
  station_col = "station_id",
  position_method = "mean",
  return_particles = TRUE,
  verbose = TRUE
)
```

# Compare with Ground Truth

```{r ground-truth}
true_tracks <- fish_simulation$tracks %>%
  rename(fish_id = path_id, time = datetime, x_true = x, y_true = y) %>%
  select(fish_id, time, x_true, y_true)

comparison <- pf_results$positions %>%
  left_join(true_tracks, by = c("fish_id", "time")) %>%
  filter(!is.na(x_true)) %>%
  mutate(error_m = sqrt((x_mean - x_true)^2 + (y_mean - y_true)^2))

cat("Mean error:", round(mean(comparison$error_m, na.rm = TRUE), 1), "m\n")
cat("Median error:", round(median(comparison$error_m, na.rm = TRUE), 1), "m\n")
cat("95th percentile:", round(quantile(comparison$error_m, 0.95, na.rm = TRUE), 1), "m\n")

comparison %>%
  group_by(fish_id) %>%
  summarise(mean_error = mean(error_m, na.rm = TRUE),
            median_error = median(error_m, na.rm = TRUE),
            n_steps = n(), .groups = "drop")
```

# Visualization

```{r plot-setup, include=FALSE}
raster_df <- as.data.frame(depth_raster, xy = TRUE)

station_coords <- sf::st_coordinates(points_regular)
stations_df <- data.frame(
  station_id = points_regular$station_id,
  station_x = station_coords[, 1],
  station_y = station_coords[, 2]
)

det_summary_by_fish <- fish_simulation$station_detections %>%
  filter(detected == 1) %>%
  rename(fish_id = path_id) %>%
  group_by(fish_id, station_id) %>%
  summarise(total_dets = n(), .groups = "drop")

fish_ids_plot <- unique(pf_results$positions$fish_id)
stations_plot <- do.call(rbind, lapply(fish_ids_plot, function(fid) {
  s <- stations_df; s$fish_id <- fid; s
}))
stations_plot <- stations_plot %>%
  left_join(det_summary_by_fish, by = c("fish_id", "station_id")) %>%
  mutate(total_dets = ifelse(is.na(total_dets), 0, total_dets),
         has_detections = total_dets > 0)

all_x <- c(comparison$x_true, comparison$x_mean)
all_y <- c(comparison$y_true, comparison$y_mean)
x_range <- range(all_x, na.rm = TRUE)
y_range <- range(all_y, na.rm = TRUE)
x_buf <- diff(x_range) * 0.15; y_buf <- diff(y_range) * 0.15
zoom_xlim <- c(x_range[1] - x_buf, x_range[2] + x_buf)
zoom_ylim <- c(y_range[1] - y_buf, y_range[2] + y_buf)
```

## Particle Cloud and Resolved Paths

```{r particle-cloud, fig.height=10}
pf_times <- unique(pf_results$particles$time)
sampled_times <- pf_times[seq(1, length(pf_times), by = 10)]
particles_sub <- pf_results$particles %>%
  filter(time %in% sampled_times) %>%
  group_by(fish_id, time) %>%
  slice_sample(n = 500) %>%
  ungroup()

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_point(data = particles_sub, aes(x = x, y = y),
             colour = "yellow", alpha = 0.1, size = 0.3) +
  geom_path(data = comparison, aes(x = x_mean, y = y_mean),
            colour = "white", linewidth = 0.5, alpha = 0.8) +
  geom_point(data = comparison, aes(x = x_mean, y = y_mean),
             colour = "white", size = 0.5, alpha = 0.6) +
  geom_path(data = comparison, aes(x = x_true, y = y_true),
            colour = "green3", linewidth = 0.4, alpha = 0.7) +
  geom_point(data = stations_plot %>% filter(!has_detections),
             aes(x = station_x, y = station_y),
             colour = "red", size = 1, shape = 4) +
  geom_point(data = stations_plot %>% filter(has_detections),
             aes(x = station_x, y = station_y, size = total_dets),
             colour = "orange", fill = NA, shape = 21, stroke = 0.8) +
  scale_size_continuous(range = c(1, 5), name = "Detections") +
  facet_wrap(~fish_id, ncol = 2) +
  coord_sf(xlim = zoom_xlim, ylim = zoom_ylim) +
  theme_minimal() +
  labs(title = "Particle Filter: Particle Cloud + Resolved Paths",
       subtitle = "Yellow = particles | White = estimated path | Green = true path")
```

## Position Error and Uncertainty

```{r error-uncertainty, fig.height=10}
ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_point(data = comparison, aes(x = x_mean, y = y_mean, colour = error_m), size = 1) +
  scale_colour_viridis_c(option = "magma", name = "Error (m)") +
  geom_segment(data = comparison,
               aes(x = x_mean - 2*x_sd, xend = x_mean + 2*x_sd,
                   y = y_mean, yend = y_mean),
               colour = "red", alpha = 0.3, linewidth = 0.2) +
  geom_segment(data = comparison,
               aes(x = x_mean, xend = x_mean,
                   y = y_mean - 2*y_sd, yend = y_mean + 2*y_sd),
               colour = "red", alpha = 0.3, linewidth = 0.2) +
  geom_path(data = comparison, aes(x = x_true, y = y_true),
            colour = "green3", linewidth = 0.3, alpha = 0.6) +
  geom_point(data = stations_plot, aes(x = station_x, y = station_y),
             colour = "grey80", size = 0.5, shape = 3) +
  facet_wrap(~fish_id, ncol = 2) +
  coord_sf(xlim = zoom_xlim, ylim = zoom_ylim) +
  theme_minimal() +
  labs(title = "Position Estimates with Error and Uncertainty",
       subtitle = "Green = true track | Red bars = 95% CI | Colour = position error")
```

## Particle Distributions at Selected Time Steps

```{r particle-snapshots, fig.height=8}
fish1_times <- comparison %>%
  filter(fish_id == comparison$fish_id[1]) %>% arrange(time)
time_idx <- round(seq(1, nrow(fish1_times), length.out = 6))
selected_times <- fish1_times$time[time_idx]

particles_snapshot <- pf_results$particles %>%
  filter(fish_id == comparison$fish_id[1], time %in% selected_times)
positions_snapshot <- comparison %>%
  filter(fish_id == comparison$fish_id[1], time %in% selected_times)

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_point(data = particles_snapshot,
             aes(x = x, y = y, colour = weight), size = 0.5, alpha = 0.4) +
  scale_colour_viridis_c(option = "magma", name = "Weight") +
  geom_point(data = positions_snapshot, aes(x = x_mean, y = y_mean),
             colour = "white", size = 2, shape = 4, stroke = 1.5) +
  geom_point(data = positions_snapshot, aes(x = x_true, y = y_true),
             colour = "green3", size = 2, shape = 3, stroke = 1.5) +
  geom_point(data = stations_plot, aes(x = station_x, y = station_y),
             colour = "grey80", size = 0.5, shape = 3) +
  facet_wrap(~time, ncol = 2) +
  coord_sf(xlim = range(c(particles_snapshot$x, positions_snapshot$x_mean), na.rm = TRUE) + c(-500, 500),
           ylim = range(c(particles_snapshot$y, positions_snapshot$y_mean), na.rm = TRUE) + c(-500, 500)) +
  theme_minimal() +
  labs(title = paste("Particle Distributions at Selected Time Steps - Fish", comparison$fish_id[1]),
       subtitle = "White cross = estimate | Green cross = true position | Colour = particle weight")
```

## Error and ESS Diagnostics

```{r diagnostics, fig.height=10}
ggplot(comparison, aes(x = time, y = error_m)) +
  geom_line(alpha = 0.5) +
  geom_point(aes(colour = n_detecting), size = 0.8) +
  scale_colour_viridis_c(name = "Detecting\nStations") +
  facet_wrap(~fish_id, scales = "free_x") +
  theme_minimal() +
  labs(title = "Position Error Over Time", x = "Time", y = "Error (m)") +

ggplot(comparison, aes(x = time, y = ess)) +
  geom_line() +
  geom_hline(yintercept = 1000 * 0.5, linetype = "dashed", colour = "red") +
  geom_point(aes(colour = resampled), size = 0.8) +
  scale_colour_manual(values = c("FALSE" = "grey50", "TRUE" = "red"), name = "Resampled") +
  facet_wrap(~fish_id, scales = "free_x") +
  theme_minimal() +
  labs(title = "Effective Sample Size Over Time",
       subtitle = "Red dashed = resampling threshold", x = "Time", y = "ESS") +

ggplot(comparison, aes(x = factor(n_detecting), y = error_m)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5) +
  theme_minimal() +
  labs(title = "Position Error by Number of Detecting Stations",
       x = "Detecting Stations", y = "Error (m)") +
  plot_layout(ncol = 1)
```

# Two-Filter Smoother

The two-filter smoother runs a backward filter from the last time step to the first, then combines forward and backward particle weights at each time step. This incorporates both past and future observations, producing more accurate position estimates.

```{r smoother}
smooth_results <- pf_smooth(
  pf_results = pf_results,
  detection_data = fish_simulation$station_detections,
  station_info = points_regular,
  de_model = logistic_DE$log_model,
  raster = depth_raster,
  step_length_mean = 50, step_length_sd = 30,
  turning_angle_sd = 45, time_step = 180,
  max_distance = 30000,
  fish_id_col = "path_id", time_col = "datetime", station_col = "station_id",
  position_method = "mean",
  return_particles = TRUE,
  verbose = TRUE
)

smooth_comparison <- smooth_results$positions %>%
  left_join(true_tracks, by = c("fish_id", "time")) %>%
  filter(!is.na(x_true)) %>%
  mutate(error_m = sqrt((x_mean - x_true)^2 + (y_mean - y_true)^2))

cat("Forward  - mean:", round(mean(comparison$error_m, na.rm = TRUE), 1),
    "m, median:", round(median(comparison$error_m, na.rm = TRUE), 1), "m\n")
cat("Smoothed - mean:", round(mean(smooth_comparison$error_m, na.rm = TRUE), 1),
    "m, median:", round(median(smooth_comparison$error_m, na.rm = TRUE), 1), "m\n")
cat("Improvement:", round((1 - mean(smooth_comparison$error_m, na.rm = TRUE) /
    mean(comparison$error_m, na.rm = TRUE)) * 100, 1), "%\n")
```

## Forward vs Smoothed Comparison

```{r smoothed-comparison, fig.height=10}
ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_path(data = comparison, aes(x = x_true, y = y_true),
            colour = "green3", linewidth = 0.4, alpha = 0.7) +
  geom_path(data = smooth_comparison, aes(x = x_mean, y = y_mean),
            colour = "white", linewidth = 0.5, alpha = 0.8) +
  geom_path(data = comparison, aes(x = x_mean, y = y_mean),
            colour = "red", linewidth = 0.3, alpha = 0.5) +
  geom_point(data = stations_plot %>% filter(!has_detections),
             aes(x = station_x, y = station_y),
             colour = "red", size = 1, shape = 4) +
  geom_point(data = stations_plot %>% filter(has_detections),
             aes(x = station_x, y = station_y, size = total_dets),
             colour = "orange", fill = NA, shape = 21, stroke = 0.8) +
  scale_size_continuous(range = c(1, 5), name = "Detections") +
  facet_wrap(~fish_id, ncol = 2) +
  coord_sf(xlim = zoom_xlim, ylim = zoom_ylim) +
  theme_minimal() +
  labs(title = "Forward vs Smoothed Position Estimates",
       subtitle = "Green = true | Red = forward | White = smoothed")
```

# Utilization Distributions

Convert particle clouds into probability-of-use surfaces and extract home range contours.

```{r utilization-distribution}
ud_results <- pf_utilization_distribution(
  pf_results = smooth_results,
  raster = depth_raster,
  contour_levels = c(0.5, 0.95),
  by_time = FALSE,
  verbose = TRUE
)

ud_results$home_ranges
```

## UD Heatmap

```{r ud-heatmap}
first_fish_id <- as.character(unique(smooth_results$positions$fish_id)[1])
ud_raster_1 <- ud_results$ud_rasters[[first_fish_id]]
ud_df <- as.data.frame(ud_raster_1, xy = TRUE)
names(ud_df)[3] <- "prob"

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  new_scale_fill() +
  geom_raster(data = ud_df %>% filter(!is.na(prob) & prob > 0),
              aes(x = x, y = y, fill = prob), alpha = 0.8) +
  scale_fill_viridis_c(option = "magma", name = "P(use)") +
  geom_path(data = comparison %>% filter(fish_id == as.integer(first_fish_id)),
            aes(x = x_true, y = y_true),
            colour = "green3", linewidth = 0.4, alpha = 0.7) +
  coord_sf(xlim = zoom_xlim, ylim = zoom_ylim) +
  theme_minimal() +
  labs(title = paste("Utilization Distribution - Fish", first_fish_id),
       subtitle = "Green = true track | Colour = probability of use")
```

## 50% and 95% Home Range Contours

```{r ud-contours, fig.height=10}
ud_contour_list <- list()
for (fid in names(ud_results$ud_rasters)) {
  ud_r <- ud_results$ud_rasters[[fid]]
  ud_vals <- as.data.frame(ud_r, xy = TRUE)
  names(ud_vals)[3] <- "prob"
  ud_vals <- ud_vals %>% filter(!is.na(prob) & prob > 0) %>%
    arrange(desc(prob)) %>%
    mutate(cum_prob = cumsum(prob),
           contour = case_when(
             cum_prob <= 0.50 ~ "Core (50%)",
             cum_prob <= 0.95 ~ "Home range (95%)",
             TRUE ~ NA_character_),
           fish_id = as.integer(fid))
  ud_contour_list[[fid]] <- ud_vals %>% filter(!is.na(contour))
}
ud_contours_df <- do.call(rbind, ud_contour_list)

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  new_scale_fill() +
  geom_raster(data = ud_contours_df,
              aes(x = x, y = y, fill = contour), alpha = 0.7) +
  scale_fill_manual(values = c("Core (50%)" = "red", "Home range (95%)" = "orange"),
                    name = "UD Contour") +
  geom_path(data = comparison, aes(x = x_true, y = y_true),
            colour = "green3", linewidth = 0.3, alpha = 0.7) +
  geom_path(data = smooth_comparison, aes(x = x_mean, y = y_mean),
            colour = "white", linewidth = 0.3, alpha = 0.5) +
  facet_wrap(~fish_id, ncol = 2) +
  coord_sf(xlim = zoom_xlim, ylim = zoom_ylim) +
  theme_minimal() +
  labs(title = "Utilization Distribution Contours (50% and 95%)",
       subtitle = "Green = true track | White = smoothed estimate")
```

# Detection Gap Analysis

Analyse how position error accumulates during periods without detections.

```{r gap-analysis}
gap_results <- pf_analyze_gaps(
  pf_results = smooth_results,
  true_tracks = fish_simulation$tracks %>%
    rename(fish_id = path_id, time = datetime, x_true = x, y_true = y) %>%
    select(fish_id, time, x_true, y_true)
)

cat("Total gaps identified:", nrow(gap_results$gaps), "\n")
if (nrow(gap_results$gaps) > 0) {
  cat("Gap duration range:",
      round(min(gap_results$gaps$gap_duration_min), 1), "-",
      round(max(gap_results$gaps$gap_duration_min), 1), "min\n")
  cat("Mean max error during gaps:",
      round(mean(gap_results$gaps$max_error_in_gap, na.rm = TRUE), 1), "m\n")
}
```

## Error vs Time Since Last Detection

```{r gap-error-scatter}
ts <- gap_results$time_series
ggplot(ts, aes(x = time_since_last_detection_sec / 60, y = error_m)) +
  geom_point(alpha = 0.3, size = 0.5, colour = "steelblue") +
  geom_smooth(method = "loess", colour = "red", se = TRUE, linewidth = 0.8) +
  theme_minimal() +
  labs(title = "Position Error vs Time Since Last Detection",
       x = "Time since last detection (min)", y = "Error (m)")
```

## Error Growth During Gaps

```{r gap-trajectories}
if (nrow(gap_results$gaps) > 0) {
  gap_ts <- ts %>% filter(!is.na(gap_id)) %>%
    left_join(gap_results$gaps %>% select(gap_id, gap_duration_min), by = "gap_id")

  ggplot(gap_ts, aes(x = time_since_last_detection_sec / 60,
                      y = error_m, group = gap_id,
                      colour = gap_duration_min)) +
    geom_line(alpha = 0.5, linewidth = 0.4) +
    scale_colour_viridis_c(name = "Gap\nduration\n(min)") +
    theme_minimal() +
    labs(title = "Error Growth During Detection Gaps",
         subtitle = "Each line = one gap event",
         x = "Time since last detection (min)", y = "Error (m)")
}
```

# Benchmark

```{r benchmark}
cat("Particle count scaling:\n")
for (np in c(100, 500, 1000, 2000)) {
  t <- system.time({
    particle_filter_positioning(
      detection_data = fish_simulation$station_detections,
      station_info = points_regular,
      de_model = logistic_DE$log_model,
      raster = depth_raster,
      n_particles = np,
      step_length_mean = 50, step_length_sd = 30,
      turning_angle_sd = 45, time_step = 180,
      max_distance = 30000,
      fish_id_col = "path_id", time_col = "datetime",
      station_col = "station_id",
      position_method = "mean",
      verbose = FALSE
    )
  })
  cat(sprintf("  %4d particles: %.1f s\n", np, t["elapsed"]))
}
```
