---
title: "🐟 Weighted Average Detection Efficiency (WADE) Positioning Analysis"
subtitle: "Positioning Animals with Acoustic Telemetry Using Detection Range Information - Simulation Application"
author: 
  - "Jake Brownscombe"
date: "`r Sys.Date()`"    
output: 
  html_document:
    theme: cosmo
    highlight: tango
    toc: true
    toc_float: 
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    code_folding: show
    fig_width: 10
    fig_height: 8
    self_contained: true
vignette: >
  %\VignetteIndexEntry{WADE Positioning Analysis}
  %\VignetteEngine{knitr::rmarkdown} 
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, 
  warning = FALSE, 
  message = FALSE,
  fig.align = 'center',
  cache = FALSE
)
```

<div class="alert alert-info">
<h4>**Analysis Overview**</h4>
This tutorial demonstrates the **Weighted Average Detection Efficiency (WADE)** approach for animal positioning using passive acoustic telemetry. It uses detection range information to generate estimates of probable animal locations based on detection patterns, which is useful for estimating space use (home range) and habitat selection. This approach using simulated data is a useful exercise to assess positioning accuracy and optimal settings relative to known tracks, which may be useful to conduct prior to field data applications where positioning can be influenced by system design, detection range, and animal behaviour. 

</div>

---

# 🚀 **Setup & Data Preparation**

## Load Required Libraries

```{r libraries}
# Core libraries for WADE positioning analysis
library(positionR)
library(raster)
library(sf)
library(ggplot2)
library(dplyr)
library(patchwork)
library(randomForest)

# Set seed for reproducible results
set.seed(123)

```

## Load Environmental Data

```{r env-data}
# Load depth raster data
data("depth_raster")
raster::crs(depth_raster) <- "EPSG:32617"  # Functions developed using UTM

```

---

# 📡 **Receiver Array Design**

## Generate Optimal Array Configuration

```{r receiver-array, fig.width=10, fig.height=8}
# Generate regularly spaced receiver array
# Note: actual number may vary slightly from n_points to optimize spacing
points_regular <- generate_exact_regular_points(
  depth_raster, 
  n_points = 80, 
  seed = 123
)

# Visualize array design
plot_points_on_input(depth_raster, points_regular) +
  ggtitle("Receiver Array Design",
          subtitle = paste("Regular spacing with", nrow(points_regular), "receivers")) +
  theme_minimal()

```

## Calculate Station Distances

```{r station-distances}
# Calculate cost-weighted distances between all grid cells and receivers
station_distances <- calculate_station_distances(
  raster = depth_raster,
  receiver_frame = points_regular,
  max_distance = 30000,  # Maximum distance to consider (30 km)
  station_col = "station_id"
)

str(station_distances) #contains shortest path, euclidean distances, tortuosity, and barrier info
```

The `calculate_station_distances()` function automatically computes barrier information (whether line-of-sight crosses land) that can be used to prevent unrealistic detections through obstacles.

---

# 📊 **Detection Range Modeling**

## Create Depth-Dependent Detection Model

```{r detection-model, fig.width=8, fig.height=8}
# Generate logistic detection efficiency model with depth-dependency
logistic_DE <- create_logistic_curve_depth(
  min_depth = 1,
  max_depth = 35,
  d50_min_depth = 400,   # 50% detection at 400m in shallow water
  d95_min_depth = 800,   # 95% detection at 800m in shallow water
  d50_max_depth = 750,   # 50% detection at 750m in deep water
  d95_max_depth = 1500,  # 95% detection at 1500m in deep water
  plot = TRUE,
  return_model = TRUE,
  return_object = TRUE
)

```

## Apply Detection Model to System

```{r apply-detection-model, fig.width=8, fig.height=6}
# Predict detection efficiency for all receiver-cell combinations
station_distances$DE_pred <- stats::predict(
  logistic_DE$log_model,
  newdata = station_distances %>%
    dplyr::rename(dist_m = cost_distance) %>%
    dplyr::mutate(depth_m = abs(raster_value)),
  type = "response"
)

# Visualize detection field for example station
ggplot(station_distances %>% dplyr::filter(station_no == 10),
       aes(x, y, fill = DE_pred)) +
  geom_raster() +
  scale_fill_viridis_c(option = "magma", name = "Detection\nEfficiency") +
  ggtitle("Detection Efficiency Field (Station 10)",
          subtitle = "Probability of detection decreases with distance and depth") +
  theme_minimal() + 
  coord_sf()

```

## System-Wide Detection Coverage

```{r system-detection, fig.width=10, fig.height=8}
# Calculate cumulative system detection efficiency
# This will be used for generating absence points later
# include_barriers = TRUE incorporates barrier masking
system_DE <- calculate_detection_system(
  distance_frame = station_distances,
  receiver_frame = points_regular,
  model = logistic_DE$log_model,
  output_type = "cumulative",  # Cumulative probability across all receivers
  plots = TRUE,
  include_barriers = TRUE  # Prevent detections through land obstacles
)

```

---

# 🐟 **Fish Movement Simulation**

## Generate Realistic Fish Tracks

```{r fish-simulation}
# Set start time for maintaining temporal structure
start_time <- as.POSIXct("2025-07-15 12:00:00", tz = "UTC")

# Simulate correlated random walks with detection events
fish_simulation <- simulate_fish_tracks(
  raster = depth_raster,
  station_distances = station_distances,  # Contains DE_pred values
  n_paths = 5,                # Number of fish to simulate
  n_steps = 480,              # Steps per fish (480 * 3 min = 24 hours)
  step_length_mean = 50,      # Mean step length (meters)
  step_length_sd = 30,        # Step length standard deviation
  time_step = 180,            # Time between steps (seconds)
  seed = 123,
  start_time = start_time,
  include_barriers = TRUE     # Prevent detections through land
)

```

## Visualize Tracks and Detections

```{r track-visualization, fig.width=12, fig.height=8}
# Plot fish tracks with detection events
plot_fish_tracks(
  fish_simulation,
  depth_raster,
  points_regular,
  show_detections = TRUE
) +
  ggtitle("Simulated Fish Tracks with Detection Events",
          subtitle = "Lines = movement paths, Points = detections, Triangles = receivers") +
  theme_minimal()
```

## Analyze Detection Performance

```{r detection-performance, fig.width=12, fig.height=8}
# Comprehensive detection performance analysis
detection_performance <- analyze_detection_performance(fish_simulation,
                                                       create_plots = TRUE,
                                                       display_plots = FALSE)

# Display performance plots in grid

performance_plots <- (detection_performance$plots$by_path + 
                      detection_performance$plots$by_station) /
                     (detection_performance$plots$distribution + 
                      detection_performance$plots$time_series)

performance_plots + 
  plot_annotation(
    title = "Detection Performance Analysis",
    subtitle = "Spatial and temporal patterns in detection probability"
  )

```

---

# 🎯 **WADE Position Estimation**

<div class="alert alert-info">
**Barrier Masking in WADE**: The barrier masking feature prevents position estimates in areas separated from receivers by land obstacles. Both detection and non-detection probabilities account for barriers, resulting in more accurate position estimates that respect physical geography.
</div>

## Calculate Position Probabilities

```{r positioning-calculation}
# Apply WADE algorithm to estimate fish positions
positioning_results <- calculate_fish_positions(
  station_detections = fish_simulation$station_detections,
  station_distances_df = station_distances,
  station_info = points_regular,
  de_model = logistic_DE$log_model,
  non_detection_weight = 0.5,          # Strength of non-detection penalty (0-1)
  integration_method = "subtractive",  # "subtractive", "multiplicative", or "additive"
  max_non_detection_distance = 2000,   # Maximum range for non-detection inference
  weighting_method = "normalize_stations",  # Normalize across stations
  # weighting_method = "information_theoretic",  # Alternative option
  normalization_method = "min_max",    # Min-max normalization
  fish_id_col = "path_id",
  time_col = "datetime",
  time_aggregation = "day",            # Daily position estimates
  station_col = "station_id",
  include_barriers = TRUE,             # Account for barriers in positioning
  verbose = TRUE
)

```

## Visualize Position Estimates

```{r position-visualization, fig.width=10, fig.height=10}
# Select example fish and time for visualization
fish_select <- 1
time_select <- "2025-07-15"

# Create three-panel plot showing different probability surfaces
p1 <- plot_fish_positions(
  positioning_results = positioning_results,
  depth_raster_df = depth_raster,
  fish_select = fish_select,
  time_select = time_select,
  plot_type = "detection",
  prob_threshold = 0.01,  # Show cells above 1% probability
  track_data = fish_simulation$tracks
) + ggtitle("Detection-Based Positioning",
           subtitle = "Probability surface from detected stations")

p2 <- plot_fish_positions(
  positioning_results = positioning_results,
  depth_raster_df = depth_raster,
  fish_select = fish_select,
  time_select = time_select,
  plot_type = "non_detection",
  track_data = fish_simulation$tracks
) + ggtitle("Non-Detection-Based Positioning",
           subtitle = "Inference from silent stations")

p3 <- plot_fish_positions(
  positioning_results = positioning_results,
  depth_raster_df = depth_raster,
  fish_select = fish_select,
  time_select = time_select,
  plot_type = "integrated",
  prob_threshold = 0.01,
  detection_threshold = 0.01,  # Filter to detection area
  track_data = fish_simulation$tracks
) + ggtitle("Integrated Position Estimate",
           subtitle = "Combined detection and non-detection evidence")

# Display all three plots
p1 + p2 + p3 + plot_spacer() + plot_layout(ncol=2)

```

---

# 🏡 **Space Use Analysis**

## Generate Position Points for Analysis

```{r position-points-generation}
# Sample points from position probability surfaces
# Initially without threshold for optimization analysis
position_points <- sample_points_from_probabilities(
  positioning_results,
  prob_column = "weighted_mean_DE_normalized_scaled",
  n_points = 1000,
  min_prob_threshold = 0.0,  # No threshold initially
  crs = 32617,
  by_group = TRUE,  # Sample per fish-time combination
  seed = 456
)

# Extract coordinates from sf object
position_points_coords <- position_points %>%
  dplyr::mutate(
    coords = sf::st_coordinates(geometry),
    x = coords[,1], 
    y = coords[,2]
  ) %>%
  sf::st_drop_geometry() %>%
  dplyr::select(-coords)

```

## Initial Space Use Estimates

```{r initial-space-use}
options(positionR.verbose = FALSE)

# Calculate positioning-based space use with initial threshold
position_space_use_results <- calculate_space_use(
  track_data = position_points_coords,
  prob_column = "probability",
  prob_thresholds = 0.2,  # Initial test threshold
  by_fish = TRUE,
  by_time_period = TRUE,
  time_aggregation = "day",
  methods = c("constrained_convex_hull"),
  grid_resolution = 100,
  reference_raster = depth_raster
)

# Calculate actual track-based space use (ground truth)
track_space_use_results <- calculate_space_use(
  track_data = fish_simulation$tracks,
  by_fish = TRUE,
  by_time_period = TRUE,
  time_aggregation = "day",
  methods = c("grid_cell_count", "constrained_convex_hull"),
  grid_resolution = 100,
  reference_raster = depth_raster
)

```

## Visual Comparison of Space Use

```{r space-use-visualization, fig.width=12, fig.height=6}
# Compare positioning vs track space use
fish_select <- 1
time_select <- "2025-07-15"

p1 <- plot_space_use(
  position_space_use_results,
  plot_type = "map",
  fish_select = fish_select,
  time_select = time_select,
  method_select = "constrained_convex_hull",
  track_data = fish_simulation$tracks,
  background_raster = depth_raster
) + ggtitle("Constrained Convex Hull Space Use (Positioning)")

p2 <- plot_space_use(
  track_space_use_results,
  plot_type = "map",
  track_data = fish_simulation$tracks,
  fish_select = fish_select,
  time_select = time_select,
  method_select = "constrained_convex_hull",
  background_raster = depth_raster
) + ggtitle("Constrained Convex Hull Space Use (Track)",
           subtitle = "Ground truth from actual movement")

p1 | p2
```

---

# 🎚️ **Probability Threshold Optimization**

## Multi-Threshold Analysis

```{r threshold-optimization, fig.width=6, fig.height=6}
options(positionR.verbose = FALSE)

# Test multiple probability thresholds
multi_results <- calculate_space_use(
  track_data = position_points_coords,
  prob_column = "probability",
  by_fish = TRUE,
  by_time_period = TRUE,
  prob_thresholds = c(0, 0.1, 0.2, 0.3, 0.4, 0.5, 0.6),
  methods = c("constrained_convex_hull"),
  reference_raster = depth_raster
)

# Compare against track-based space use
# Note: Track overlap calculation may have issues
comparison <- compare_space_use_thresholds(
  multi_threshold_results = multi_results,
  reference_tracks = fish_simulation$tracks,
  reference_space_use = track_space_use_results
)

# Display optimization results
print(comparison$plots$area_ratio_boxplot) 
print(comparison$plots$track_overlap_boxplot)

```

## Final Space Use with Optimal Threshold

```{r final-space-use, fig.width=8, fig.height=6}
# Calculate space use with optimized threshold
space_use_0.4 <- calculate_space_use(
  track_data = position_points_coords,
  prob_column = "probability",
  by_fish = TRUE,
  by_time_period = TRUE,
  prob_thresholds = 0.4,  # Optimal threshold from analysis
  methods = c("constrained_convex_hull"),
  reference_raster = depth_raster
)

# Visualize final space use estimates
ggplot(space_use_0.4$space_use_estimates, 
       aes(as.factor(fish_id), constrained_convex_hull_area_hectares)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  geom_jitter(width = 0.1, alpha = 0.5, size = 2) +
  theme_minimal() +
  labs(
    title = "Positioning Estimates of Space Use",
    subtitle = "Using optimized probability threshold (0.4)",
    x = "Fish ID",
    y = "Space Use (hectares)"
  )

# Summary statistics
space_summary <- space_use_0.4$space_use_estimates %>%
  group_by(fish_id) %>%
  summarise(
    mean_area = mean(constrained_convex_hull_area_hectares),
    sd_area = sd(constrained_convex_hull_area_hectares),
    .groups = 'drop'
  )

```

---

# 🌿 **Habitat Selection Analysis**

## Generate Presence Points for Habitat Analysis

```{r presence-points-habitat, fig.width=10, fig.height=8}
# Note: Need to integrate filtering approaches from field applications
# Generate presence points using optimal threshold
position_points <- sample_points_from_probabilities(
  positioning_results,
  prob_column = "weighted_mean_DE_normalized_scaled",
  n_points = 1000,
  min_prob_threshold = 0.4,  # Based on optimization results
  crs = 32617,
  by_group = TRUE,
  seed = 456
)

# Visualize presence points
raster_df <- as.data.frame(depth_raster, xy = TRUE)

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_sf(data = position_points %>% 
            filter(fish_id == 1 & time_period_label == "2025-07-15"), 
          aes(color = probability),
          alpha = 0.5, size = 1) +
  scale_color_viridis_c(name = "Probability", option = "magma") +
  theme_minimal() +
  labs(title = "Presence Points",
       subtitle = "Fish 1, July 15 - Points weighted by position probability")

```

## Generate Absence Points

### System DE-Weighted Absences

```{r absence-points-weighted, fig.width=10, fig.height=8}
# Generate absences weighted by system detection efficiency
absence_points <- sample_points_from_system_de(
  system_DE,
  position_points = position_points,
  n_points = 1000,
  min_prob_threshold = 0.05,
  crs = 32617,
  seed = 123
)

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_sf(data = absence_points %>% 
            filter(fish_id == 1 & time_period_label == "2025-07-15"), 
          aes(color = probability),
          alpha = 1, size = 0.8) +
  scale_color_viridis_c(name = "Cumulative\nProbability") +
  theme_minimal() +
  labs(title = "Absence Points (System DE Weighted)",
       subtitle = "Sampling probability proportional to detection coverage")
```

### Uniform Random Absences

```{r absence-points-uniform, fig.width=10, fig.height=8}
# Generate uniform absences across entire system
absence_points_uniform <- sample_points_from_system_de(
  system_DE,
  position_points = position_points,
  n_points = 1000,
  uniform = TRUE,  # Uniform sampling
  min_prob_threshold = 0.0,  # No threshold - whole system
  crs = 32617,
  seed = 123
)

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_sf(data = absence_points_uniform %>% 
            filter(fish_id == 1 & time_period_label == "2025-07-15"),
          alpha = 1, size = 0.8, col = "red") +
  theme_minimal() +
  labs(title = "Absence Points (Uniform Across System)",
       subtitle = "Random sampling across entire study area")

```

## Depth Selection Analysis - Positioning

```{r depth-selection-positioning, fig.width=12, fig.height=6}
# Extract depth values
position_points$depth_m <- raster::extract(depth_raster, position_points)
absence_points_uniform$depth_m <- raster::extract(depth_raster, absence_points_uniform)

# Combine presence and absence data
# Using uniform absences for consistency with track analysis
presence_absence_points <- rbind(
  position_points %>%
    select(fish_id, time_period_posix, depth_m) %>%
    mutate(type = "presence"),
  absence_points_uniform %>%
    select(fish_id, time_period_posix, depth_m) %>%
    mutate(type = "absence")
)

# Create depth selection plots
p1 <- plot_depth_selection(presence_absence_points, plot_type = "density") +
  ggtitle("Depth Selection",
          subtitle = "Blue = presences, Red = absences")

p2 <- plot_depth_comparison(
  presence_absence_points,
  comparison_var = "fish_id",
  plot_type = "boxplot"
) +
  ggtitle("Depth Selection by Individual Fish")

p1 | p2


```

---

# 🐠 **Track-Based Habitat Selection (Ground Truth)**

## Generate Track Presence Points

```{r track-presence, fig.width=10, fig.height=8}
# Sample points from grid cells used by actual tracks
track_presence <- sample_points_from_track_grid(
  track_data = fish_simulation$tracks,
  reference_raster = depth_raster,
  n_points = 1000,
  crs = 32617
)

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_sf(data = track_presence, 
          aes(color = count),
          alpha = 0.5, size = 1) +
  scale_color_viridis_c(name = "Count", option = "magma") +
  coord_sf() +
  theme_minimal() +
  labs(title = "Track Presence Points",
       subtitle = "Sampled from actual movement paths")
```

## Generate Track Absence Points

```{r track-absence, fig.width=10, fig.height=8}
# Generate absences for track-based analysis
track_absence <- sample_points_from_system_de(
  system_DE,
  position_points = track_presence,
  n_points = 1000,
  uniform = TRUE,  # Uniform sampling
  min_prob_threshold = 0.0,  # Whole system
  crs = 32617,
  seed = 123
)

ggplot() +
  geom_raster(data = raster_df, aes(x = x, y = y, fill = layer)) +
  scale_fill_gradient(low = "blue4", high = "cornflowerblue",
                      na.value = "transparent", name = "Depth (m)") +
  geom_sf(data = track_absence %>% 
            filter(fish_id == 1 & time_period_label == "2025-07-15"),
          alpha = 1, size = 0.8, col = "red") +
  theme_minimal() +
  labs(title = "Track Absence Points (Uniform Across System)",
       subtitle = "Random background points for comparison")
```

## Depth Selection Analysis - Tracks

```{r depth-selection-tracks, fig.width=12, fig.height=6}
# Extract depth values for track data
track_presence$depth_m <- raster::extract(depth_raster, track_presence)
track_absence$depth_m <- raster::extract(depth_raster, track_absence)

# Combine track presence and absence data
track_presence_absence_points <- rbind(
  track_presence %>%
    select(fish_id, time_period_posix, depth_m) %>%
    mutate(type = "presence"),
  track_absence %>%
    select(fish_id, time_period_posix, depth_m) %>%
    mutate(type = "absence")
)

# Create track depth selection plots
p1 <- plot_depth_selection(track_presence_absence_points, plot_type = "density") +
  ggtitle("Depth Selection (Track-Based Ground Truth)",
          subtitle = "Blue = presences, Red = absences")

p2 <- plot_depth_comparison(
  track_presence_absence_points,
  comparison_var = "fish_id",
  plot_type = "boxplot"
) +
  ggtitle("Depth Selection by Individual Fish (Tracks)")

p1 | p2


```

---

# 🔬 **Comparative Habitat Selection Analysis**

## Random Forest Model Comparison

```{r comparative-analysis, fig.width=12, fig.height=10}
# Comprehensive comparison using Random Forests
comparative_results <- analyze_comparative_habitat_selection(
  positioning_data = presence_absence_points,
  space_use_data = track_presence_absence_points,
  analysis_type = "both",
  formula = presence ~ depth_m + fish_id + time_period_posix,
  time_variable = "time_period_posix",
  sample_size = 10000,
  ntree = 500,
  create_plots = TRUE,
  create_comparison = TRUE
)

# Display comprehensive summary
cat("=== COMPARATIVE HABITAT SELECTION ANALYSIS ===\n\n")
print_comparative_summary(comparative_results)
```

## Visualize Comparative Results

```{r comparative-plots, fig.width=12, fig.height=10}
# Plot all comparative results
plot_comparative_results(comparative_results, plot_type = "comparison")

```

---

# 📋 **Summary & Conclusions**

<div class="alert alert-success">
<h2>**Key Takeaways**</h2>

- Here we've used a simulation based dataset to conduct WADE positioning,
which produces useful outputs for assessing space use (home range) and
habitat selection from acoustic telemetry data. 

- There are a variety of options for the WADE algorithm in how the data 
are processed (e.g., normalizing station detection ranges, weighting of
detections vs non-detecting receivers, threshold cut offs for positioning). 
Conducting simulation-based analysis within your system can aid in determining the 
optimal approaches relative to system design, detection efficiency, and
animal behaviour by comparing the accuracy of positioning outputs with the 
known track data. 

- Tools for generating reasonable detection range models that integrate depth
are provided, although it is much more optimal to use real detection range 
data and models to inform this. Detection range varies considerably over space and
time due to variation in environmental conditions. 

</div>

